Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2023-04-21 analysis/paper/paper.Rmd
writing in progress

---
title: "Implementation and assessment of methods for projecting forest road network development"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
    equal_contributor: "yes"
    orcid_id: 0000-0001-7875-9015
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
    equal_contributor: "yes"
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

institute:
  - ECCC: Environment and Climate Change Canada 
  - FLNR: British Columbia Ministry of Forests 
  - UBC: University of British Columbia 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: Roads.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  chunk_output_type: inline
---

@@

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, very few of these projections have included road networks. For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire. This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012] and tools exist [e.g. @remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate]. However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. Hardy's tool is designed for integration with other components of the Landis-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

@@

We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land where road building is difficult or impossible [new tsa data package ref]. Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain. Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data. The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River. At low to mid-elevations, the forests of the interior cedar-hemlock biogeoclimatic zone (bec ref) include remnants of rare old growth inland temperate rainforest (parks canada website). At higher elevations, forests are dominated by Englemann spruce and subalpine fir (bec ref). The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt (new tsa data package ref). Two of six

Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife of concern including including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

<!--	TO DO: consider these comments from Kyle: Need a description about road development. In BC, forestry roads are developed by forest licensees under a timber harvesting permit. Various considerations go into road development including soils, grade, and economics. What type/classification of roads are you looking at (primitive roads, paved roads, etc). What’s the maximum grade? ---can find a lot of this verbiage in the Interior Appraisal Manual. How are they maintained – do they last forever -see Auditor General Report 
•	How much of the Revelstoke TSA overlaps with caribou range? Elevation? Climate? Flora/Fauna?
•	Maybe add a table of descriptive statistics presenting information about roads in the Revelstoke TSA and the other TSAs you mention? By Caribou Range? – road density, road length, cutblock area, cutblock density, etc

-->

@@

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems. For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design. The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing. The method was first developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings and evaluating road branching opportunities. Dean [-@dean_finding_1997] showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network. These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes. The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node with a tree that minimizes edge weights and does not include loops. This method differs from Clark et al. [@clark_three-stage_2000] who used Euclidean distances for edge weights.

@@

## Aspatial performance
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). Within all mapped cutblocks (Figure \@(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road. Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@(fig:aspat-perf-fig-all)). These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road. However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuousity within cutblocks (Figure \@ref(fig:proj-meth-fig)). None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).

@@

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results. These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022] that users should keep in mind interpreting discrepancies between observed and projected road networks.  
Outside cutblocks, we saw very little variation in performance among projection method variants. All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads. Outside cutblocks our DLCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads. Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm. The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.

@@

@@

Discuss possible other uses of algorithms - non forest roads, attributing age to segments as in lochhead, linking disconnected networks (but see coops et al refs)

@@

<!-- TO DO: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. 
<!--  What David had

Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc).

The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries. In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).

Many spatial stochastic models don't include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010). Our model could be relatively easily incorporated, or applied retroactively, to similar spatial stochastic simulations such as these. For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.
-->
```
# Acknowledgements

>>>>

---
title: "Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
    equal_contributor: "yes"
    orcid_id: 0000-0001-7875-9015
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
    equal_contributor: "yes"
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

institute:
  - ECCC: Environment and Climate Change Canada 
  - FLNR: British Columbia Ministry of Forests 
  - UBC: University of British Columbia 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: Roads.bib
csl: "../templates/canadian-journal-of-forest-research.csl" # Insert path for the bib-style
abstract: |
  abstract to do
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  chunk_output_type: inline
---

@@

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, very few of these projections have included road networks. For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire. This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g. @remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist. However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. Hardy's tool is designed for integration with other components of the Landis-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

@@

We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is difficult or impossible. Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain. Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data. The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River. At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_and_range_biogeoclimatic_nodate] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017]. At higher elevations, forests are dominated by Englemann spruce and subalpine fir [@british_columbia_ministry_of_forests_and_range_biogeoclimatic_nodate]. The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022]. Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020;@british_columbia_ministry_of_forests_revelstoke_2022;@environment_canada_recovery_2014]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

<!--	TO DO: consider these comments from Kyle: Need a description about road development. In BC, forestry roads are developed by forest licensees under a timber harvesting permit. Various considerations go into road development including soils, grade, and economics. What type/classification of roads are you looking at (primitive roads, paved roads, etc). What’s the maximum grade? ---can find a lot of this verbiage in the Interior Appraisal Manual. How are they maintained – do they last forever -see Auditor General Report 
•	How much of the Revelstoke TSA overlaps with caribou range? Elevation? Climate? Flora/Fauna?
•	Maybe add a table of descriptive statistics presenting information about roads in the Revelstoke TSA and the other TSAs you mention? By Caribou Range? – road density, road length, cutblock area, cutblock density, etc
-->

@@

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems. For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The simple underlying grid is less computationally costly than methods designed to support curvature or switchbacks required in operational road design [@stuckelberger_improved_2007]. The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing. The method was developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings and evaluating road branching opportunities. Dean [-@dean_finding_1997] showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network. These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes. The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links all nodes with a tree that minimizes edge weights and does not include loops. This method differs from Clark et al. [@clark_three-stage_2000] who used Euclidean distances for edge weights.

TO DO: talk about key differences from Hardy algorithm.
TO DO: talk about implications of algorithms for speed/computational cost.

@@

## Aspatial performance
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). Within all mapped cutblocks (Figure \@ref(fig:aspat-perf-fig-all)), the Hardy method most accurately projected road density, and least accurately projected distance to road. Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks (Figure \@ref(fig:aspat-perf-fig-all)). These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the analysis (Figure \@ref(fig:aspat-perf-fig)) clarifies that within real cutblocks the Hardy method most accurately projected road density and distance to road. However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain , and the Hardy method projects a distinctly different pattern of road tortuousity within cutblocks (Figure \@ref(fig:proj-meth-fig)). None of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).

@@

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results. These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.

Outside cutblocks, we saw very little variation in performance among projection method variants. All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads. Outside cutblocks our DLCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads. Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm. The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.

@@

In this paper our focus is tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife. However, we note there are other possible uses for these road network projection methods. Cost minimizing methods are not suitable for projecting development of seismic lines (Figure \@ref(fig:ftNelson)), but for other circumstances such as mine access where the objective is to minimize the cost of reaching a target location these methods may be applicable. Lochhead et al. [-@lochhead_linking_2022] used the least cost path algorithm to reconstruct the likely history of road network development in a case where harvest dates were known but road construction dates were not. In principle, these algorithms could also be used to link disconnected road segments into a network, but local route finding algorithms [@roussel_vectorial_2023] that do not require construction of large graphs are better suited to this purpose. More obviously, spatial stochastic models are used to project impacts of disturbance on a variety of values other than wildlife [@albrich_simulating_2020; @hof_editorial_2021; @sturtevant_understanding_2021]. It would also be interesting to account for persistent impacts of resource roads on vegetation, productivity and carbon sequestration [@braham_characterization_2023].

@@

@@

<!-- TO DO: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. 
-->
```
# Acknowledgements
